My talk covered one of the main results
in the paper:

J.-L. Colliot-Th\'el\`ene, J.-J. Sansuc et Sir Peter
Swinnerton-Dyer) Intersections of two quadrics and Ch\^atelet surfaces,
I, J. f\"ur die reine und angew. Math. (Crelle) {\bf 373} (1987) 37-107;
II, ibid. {\bf 374} (1987) 72-168.



 Other papers relevant to my talk at
the Winter School are:

 J.-L. Colliot-Th\'el\`ene et J.-J. Sansuc,  La descente sur les
vari\'et\'es rationnelles.  In Journ\'ees de g\'eom\'etrie alg\'ebrique
d'Angers (Juillet 1979),\'edit\'e par  A.Beauville, Sijthof and Noordhof
(1980) 223-237.


 J.-L. Colliot-Th\'el\`ene et J.-J. Sansuc, La descente sur les
vari\'et\'es rationnelles II,  Duke Math. J. {\bf 54} (1987) 375-492.


  J.-L. Colliot-Th\'el\`ene, Surfaces rationnelles fibr\'ees en
coniques de degr\'e 4,  S\'eminaire de  th\'eorie des nombres de Paris
88-89. Progress in Mathematics, Birkh\"auser (1990), p. 43-55.

 J.-L. Colliot-Th\'el\`ene et P. Swinnerton-Dyer, Hasse principle and
weak approximation for pencils of Severi-Brauer and similar varieties
J. f\"ur die reine und  ang. Mathematik (Crelle) {\bf 453} (1994) 49-112.

  J.-L. Colliot-Th\'el\`ene, A. N. Skorobogatov et Sir Peter
Swinnerton-Dyer,  Rational points and zero-cycles on fibred varieties :
Schinzel's hypothesis and Salberger's device,
 Journal f\"ur die reine und angewandte
Mathematik (Crelle) {\bf 495}  (1998) 1-28.

 J.-L. Colliot-Th\'el\`ene, A. N. Skorobogatov et Sir Peter
Swinnerton-Dyer,  Hasse principle  for pencils of curves of genus one
whose Jacobians have rational 2-division points,
Inventiones math. {\bf 134} (1998) 579-650.

 J.-L. Colliot-Th\'el\`ene and  A. Skorobogatov,
Descent on fibrations over ${\bf P}^1_k$ revisited, November 1998.
To appear in Mathematical Proceedings
of the Cambridge Philosophical Society (2000).




Surveys :

 J.-L. Colliot-Th\'el\`ene, Arithm\'etique des vari\'et\'es
rationnelles et probl\`emes birationnels.  Proceedings of the
International Congress of Mathematicians, Berkeley,  California 1986,
(1987), Tome I, 641-653.

 J.-L. Colliot-Th\'el\`ene, L'arithm\'etique des vari\'et\'es
rationnelles.
Annales de la Facult\'e des sciences de Toulouse, vol. {\bf I}, num\'ero
3 (1992) 295-336.

 J.-L. Colliot-Th\'el\`ene,The Hasse principle in a pencil of
algebraic varieties,
 Proceedings of the Tiruchirapalli
conference (India, January 1996), ed. M. Waldschmidt and K.
Murty, Contemporary Mathematics, {\bf 210} (1998) 19-39.



Almost all these papers have appeared, as you may see.
 Here is the address of my web page :

http://www.math.u-psud.fr/\~{}colliot/


There people will find links to preprints.

\end
