\documentclass{article}\usepackage{amssymb}\begin{document}\Large\begin{center}\underline{\bf {A History of Interactions between}}\underline{\bf{ Logic andNumber Theory}}\vspace{3mm}A. Macintyre\end{center}\normalsize\section{Outline}Interactions between logic and number theory have almost alwaysinvolved \underline{definability}.In an earlier period the relevant logical component was recursiontheory (decidability and undecidability). For ${\mathbb Z}$ thecentral issue was Hilbert's 10th Problem, and the central resultis that recursively enumerable relations on ${\mathbb Z}$ areexistentially definable. The highpoint of definability theory in${\mathbb Q}$ remains Julia Robinson's, that ${\mathbb Z}$ is$\Pi_3$-definable in ${\mathbb Q}$. Whether ${\mathbb Z}$ isexistentially definable in ${\mathbb Q}$ is unknown (if it is,Hilbert's 10th Problem for ${\mathbb Q}$ is undecidable).Recursion theory is thus very relevant for the logic of\underline {global} fields and their rings of integers.In contrast, model theory is much more relevant for the logic oflocal fields, and for those areas of number theory with ageometric aspect.The locally compact completions of number fields have allundergone  fruitful model-theoretic analyses. Thus Tarski (1930's)obtained the classical results on definitions in ${\mathbb C}$ and${\mathbb R}$, while not till the 1960's did Ax-Kochen-Ersov obtainanalogous results for p-adic fields (and for many Henselianfields). The completions are all decidable, but nowadays one givesmore importance to the definability aspect of the above analyses.One obtains normal forms for definitions, links between thegeometry of the set and the form of its definition, and variousuniformities for number of connected components, and in ${\mathbb Q}_p$ for the form of various integrals (cf.~Loeser's course).In the 1980's the ring of all algebraic integers was shown(via a local-to-global principle involving earlier work inalgebraically closed fields with valuation) to have a very cleardefinability theory, and in particular to have the analogue ofHilbert's 10th Problem decidable.In the 1960's and 70's there were several developments in\underline{pure} model theory that led some time later tointeractions with number theory. The first was the work ofRobinson school on model completion, existentially closedstructures, and forcing methods. The theories of the completions${\mathbb C}$, ${\mathbb R}$, ${\mathbb Q}_p$ all arise naturally inthis settings. But other \underline{theories} emerge, withoutnatural models, but which were to be key components insignificant interactions in the 1990's. One is the theory ofdifferentially closed fields (to be involved in the Mordell-Langconjecture, cf.~the Pillay-Scanlon course), and another (notdiscovered till 1990) is the theory ACFA of generic automorphisms(to be involved in the ``logical" approach to the Manin-Mumfordconjecture). Other theories of this type relate to the lifting ofFrobenius to the Witt vectors.The other model-theoretic development, certainly deeper qua modeltheory, originated with Morley's (1965) work initiatingmodel theoretic stability. The 1965 paper gave a suggestivetopological setting for first-order definability,and initiated asystematic study of general notions around the geometrical ideasof \underline{dimension} and \underline {independence}. Although theensuing \underline{stability theory} applies only to ${\mathbb C}$among the completions of number fields, latter day ``local" versionsof it have been involved in most recent interactions of logic andnumber theory (cf.~Pillay-Scanlon).The other turning point in the 60's was Ax's work on theelementary theory of finite fields,where logic was seen tointeract with Weil's Riemann Hypothesis for curves, and withCebotarev's Theorem. The new theories could be construed ascompletions of Robinson type, and their definability wassuggestively analyzed in terms of \underline{GaloisStratification} (cf.~Loeser's lectures). Moreover, one was soon ledto model theoretic questions about \underline{absolute Galoisgroups}, and those are related to a vision of Grothendieck (seePop's lectures). (This is by no means the only case where ideas ofGrothendieck, the ``logician" of Bourbaki, have had fundamentallogical content.)Already in Ax-Kochen-Ersov one had seen the power of the impulse ``Let $p$ go to 0."  After Ax's work one had richer environments forthis idea. In particular, one could gradually approach the studyof the Frobenius $x^p$ as $p$ tends to 0. Here one makes essentialcontact with the Weil Conjectures and the cohomological methodsused in their proof. Old themes of Robinson, on bounds in thetheory of ideals in polynomial rings, reappear in the widersetting of Intersection Theory and Weil Cohomology, and relate tothe Grothendieck Standard Conjectures.Another grand conjecture, that of Schanuel on transcendence ofvalues of the complex exponential, has recently begun to interactwith logic. It was first seen in connection with the decidabilityof the real and p-adic exponentials, and more recently in aprofound definability-theoretic study by Zilber of the\underline{undecidable} complex exponential. Zilber's workinteracts naturally with diophantine geometry, and with old work ofAx (cf.~Pillay-Scanlon). \section{Structure of lectures}\begin{enumerate}\itemDefinabilityin the fields ${\mathbb C}$, ${\mathbb R}$ and ${\mathbb Q}_{p}$.Uniformities, with specialreference on those in $p$.\itemModel theory. New theories. Ax's work.\itemFrom pseudofinite to ACFA. Galois groups and logic. Manifestations ofFrobenius. \itemAnalytic aspects. Schanuel's Conjecture.\end{enumerate}\section{Prerequisites}Basics of first-order logic, algebraic geometry, and numbertheory. (Roughly as for Poonen's course).\section{Project}The completions of number fields are naturally united inthe adele construction. 25 years ago Weispfenning gave a first analysisof definability in the adeles, in the spirit of Feferman-VaughtTheorems. In view of the deepening of our understanding of definabilityin the interim (e.g., in the work of Denef and Loeser),  it seems natural togo back and write an up-to-date account, paying attention tomeasure-theoretic and analyic uniformities. For example, what exactly dothe model theoretic uniformities for $p$-adic integrals mean in an adelicsetting? The goal is to give a Feferman-Vaught analysis for the analyticstructure of the adeles.I am currently having trouble tracking down the reference forWeispfenning's original work, but am confident of locating it beforelong.\begin{thebibliography}{10}\bibitem{fj}Fried, Michael D., and Jarden, Moshe {\em Field arithmetic}.\newblock Ergebnisse der Mathematik   und ihrer Grenzgebiete (3) [Results in Mathematics and Related Areas(3)], 11.   Springer-Verlag, 1986.\bibitem {m1}     Macintyre, Angus    {\em Twenty years of $p$-adic model theory}.    \newblock Logiccolloquium   '84 (Manchester, 1984), 121--153, Stud. Logic Found. Math., 120,North-Holland,   Amsterdam, 1986.   \bibitem{mw1}Macintyre, Angus; Wilkie, A. J. {\em On the decidability of the realexponential   field}.    \newblock Kreiseliana, 441--467, A K Peters, Wellesley, MA, 1996.   \bibitem{ma}Marker, David Model Theory:An Introduction.\newblock Graduate Texts in Mathematics217, Springer Verlag, 2002.\bibitem{hand}   {\em Handbook of mathematical logic}. Edited by Jon Barwise. With the   cooperation of H. J. Keisler, K. Kunen, Y. N. Moschovakis and A. S.Troelstra. \newblock Studies in Logic   and the Foundations of Mathematics, Vol. 90. North-Holland PublishingCo., Amsterdam-New   York-Oxford, 1977.\end{thebibliography}\end{document}