\documentstyle[12pt]{article}

\begin{document}
\parskip=9pt

\begin{center}
{\Large\bf References for\\Peter Sarnak's course\\at AWS 2001}
\end{center}


\noindent{\bf SURVEY:}

1.\ \ Iwaniec and Sarnack, ``Perspectives on the analytic theory
of $L$-functions," to appear in {\it GAFA} (2001).

\bigskip
\noindent{\bf SUBCONVEXITY:}

2.\ \ Friedlander and Iwaniec, ``A mean value theorem for
character sums," {\it Michigan Math Jnl} {\bf 39}, 153-159 (1992).

3.\ \ Duke, Friedlander and Iwaniec, ``Bounds for automorphic
$L$-functions I and II," {\it Invent. Math} {\bf 112}, 1-8 (1993);
{\it Invent. Math} {\bf 115}, 219-239 and 209-217 (1994).

4.\ \ Sarnak, ``Estimates for Rankin Selberg $L$-functions and
quantum ergodicity," preprint 2001.

5.\ \ Sarnak, ``Integrals of products of eigenfunctions," {\it
Imrn} {\bf 6}, 251-260 (1994).

\bigskip
\noindent{\bf SPECIAL VALUES OF $L$-FUNCTIONS:}

6.\ \ Waldspurger, ``Sur les coefficients de Fourier -- demi
enitier," {\it J. Math. Pures et Appl.} {\bf 60}, 365-384 (1981).

7.\ \ T. Watson, ``Rankin triple products and quantum chaos," {\it
Thesis Princeton U} (2001).

\bigskip
\noindent{\bf NON-VANISHING AND MOLLIFICATION:}

8.\ \ Iwaniec and Sarnak, ``The non-vanishing of central values of
automorphic $L$-functions and Landau-Siegel zeros," {\it Israel
Math Jnl}, Feb. 2001.

9.\ \ Kowalski, Michel and Vanderkam, ``Mollification of the
fourth moments of automorphic $L$-functions and arithmetic
applications," {\it Invent.} {\bf 142}, 95-151 (2000).

\bigskip
\noindent{\bf PETERSSON FORMULA:}

10.\ \ Iwaniec, "Topics in classical automorphic forms," {\it
G.S.M. AMS}, Vol. {\bf 17} (1997).

\bigskip
\noindent{\bf FAMILIES AND SYMMETRY:}

11.\ \ Katz and Sarnack, "Zeros of zeta functions and symmetry,"
{\it BAMS}, {\bf 36}, 1-26 (1999).

12.\ \ Keating and Snaith, "Random matrix theory and $L$-functions
at $S=\frac12$," {\it Comm. Math Phys.} {\bf 214}, 91-110 (2000).

13.\ \ P. Michel, ``Repartition des zeros des functions $L$ et
matrices aleatoires," {\it Seminar Bourbaki\/}, March 2001.



\end{document}
